---
title: "Second step: Formating"
author: "Amos Bationo"
date: "2023-04-08"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE}
library("tidyverse")
library("lubridate")
print("Packages loaded!")
```

# -------------- Read all extracted csv files
#______________________________________________________________________________#
### ---- Data sets
```{r Data sets, include=FALSE}
df_2015 <- read_csv(file.path('data','data_2015.csv'))
df_2016 <- read_csv(file.path('data','data_2016.csv'))
df_2017 <- read_csv(file.path('data','data_2017.csv'))
df_2018 <- read_csv(file.path('data','data_2018.csv'))
df_2019 <- read_csv(file.path('data','data_2019.csv'))
df_2020_Q1 <- read_csv(file.path('data','data_2020_Q1.csv'))
df_2020_Q2_to_2020_11 <- read_csv(file.path('data','data_2020_Q2_to_2020_11.csv'))
df_2020_12_to_2023 <- read_csv(file.path('data','data_2020_12_to_2023.csv'))
print("Data sets loaded!")
```

```{r}
filter(df_2020_12_to_2023, is.na(end_station_name))
```

Display the first 6 trips in each data set
```{r}
head(df_2015)
head(df_2016)
head(df_2017)
head(df_2018)
head(df_2019)
head(df_2020_Q1)
head(df_2020_Q2_to_2020_11)
head(df_2020_12_to_2023)
```
Display the last 6 trips in each data set
```{r}
tail(df_2015)
tail(df_2016)
tail(df_2017)
tail(df_2018)
tail(df_2019)
tail(df_2020_Q1)
tail(df_2020_Q2_to_2020_11)
tail(df_2020_12_to_2023)
```

# -------------- Fix columns-related issues
#______________________________________________________________________________#

Format times: We want all times to be in 'yyyy-mm-dd hh:mm:ss' format
```{r}
df_2015$starttime <- mdy_hm(df_2015$starttime)
df_2015$stoptime <- mdy_hm(df_2015$stoptime)

# df_2016 contains 2 format: mm/dd/yy hh:mm:ss and mm/dd/yy hh:mm
### starttime
df_2016_1 <- filter(df_2016, str_count(starttime, ":") == 1) # for mm/dd/yy hh:mm
df_2016_2 <- filter(df_2016, str_count(starttime, ":") == 2) # mm/dd/yy hh:mm:ss
df_2016_1$starttime <- mdy_hm(df_2016_1$starttime)
df_2016_2$starttime <- mdy_hms(df_2016_2$starttime)

df_2016 <- bind_rows(df_2016_1, df_2016_2)

### stoptime
df_2016_1 <- filter(df_2016, str_count(stoptime, ":") == 1) # for mm/dd/yy hh:mm
df_2016_2 <- filter(df_2016, str_count(stoptime, ":") == 2) # mm/dd/yy hh:mm:ss
df_2016_1$stoptime <- mdy_hm(df_2016_1$stoptime)
df_2016_2$stoptime <- mdy_hms(df_2016_2$stoptime)

df_2016 <- bind_rows(df_2016_1, df_2016_2)

# df_2017
df_2017$start_time <- mdy_hms(df_2017$start_time)
df_2017$end_time <- mdy_hms(df_2017$end_time)

# df_2018 # Already in the desired format

# df_2019 # Already in the desired format

# df_2020_Q1 # Already in the desired format

# df_2020_Q2_to_2020_11 # Already in the desired format
print("Done.")
```

Merge columns containing same information (not identical data points) from 
### different csv files
#### For 2018
```{r}
colnames(df_2018)
```
```{r}
df_2018$trip_id[is.na(df_2018$trip_id)] <- 
  df_2018$X01...Rental.Details.Rental.ID[is.na(df_2018$trip_id)]
df_2018$bikeid[is.na(df_2018$bikeid)] <- 
  df_2018$X01...Rental.Details.Bike.ID[is.na(df_2018$bikeid)]
df_2018$start_time[is.na(df_2018$start_time)] <- 
  df_2018$X01...Rental.Details.Local.Start.Time[is.na(df_2018$start_time)]
df_2018$end_time[is.na(df_2018$end_time)] <- 
  df_2018$X01...Rental.Details.Local.End.Time[is.na(df_2018$end_time)]
df_2018$tripduration[is.na(df_2018$tripduration)] <- 
  df_2018$X01...Rental.Details.Duration.In.Seconds.Uncapped[is.na(df_2018$tripduration)]
df_2018$from_station_name[is.na(df_2018$from_station_name)] <- 
  df_2018$X03...Rental.Start.Station.Name[is.na(df_2018$from_station_name)]
df_2018$to_station_name[is.na(df_2018$to_station_name)] <- 
  df_2018$X02...Rental.End.Station.Name[is.na(df_2018$to_station_name)]
df_2018$from_station_id[is.na(df_2018$from_station_id)] <- 
  df_2018$X03...Rental.Start.Station.ID[is.na(df_2018$from_station_id)]
df_2018$to_station_id[is.na(df_2018$to_station_id)] <- 
  df_2018$X02...Rental.End.Station.ID[is.na(df_2018$to_station_id)]
df_2018$usertype[is.na(df_2018$usertype)] <- 
  df_2018$User.Type[is.na(df_2018$usertype)]
print("Done.")
```

```{r}
head(filter(df_2018, is.na(start_time)))
```


#### For 2019
```{r}
colnames(df_2019)
```

```{r}
df_2019$trip_id[is.na(df_2019$trip_id)] <- 
  df_2019$X01...Rental.Details.Rental.ID[is.na(df_2019$trip_id)]
df_2019$bikeid[is.na(df_2019$bikeid)] <- 
  df_2019$X01...Rental.Details.Bike.ID[is.na(df_2019$bikeid)]
df_2019$start_time[is.na(df_2019$start_time)] <- 
  df_2019$X01...Rental.Details.Local.Start.Time[is.na(df_2019$start_time)]
df_2019$end_time[is.na(df_2019$end_time)] <- 
  df_2019$X01...Rental.Details.Local.End.Time[is.na(df_2019$end_time)]
df_2019$tripduration[is.na(df_2019$tripduration)] <- 
  df_2019$X01...Rental.Details.Duration.In.Seconds.Uncapped[is.na(df_2019$tripduration)]
df_2019$from_station_name[is.na(df_2019$from_station_name)] <- 
  df_2019$X03...Rental.Start.Station.Name[is.na(df_2019$from_station_name)]
df_2019$to_station_name[is.na(df_2019$to_station_name)] <- 
  df_2019$X02...Rental.End.Station.Name[is.na(df_2019$to_station_name)]
df_2019$from_station_id[is.na(df_2019$from_station_id)] <- 
  df_2019$X03...Rental.Start.Station.ID[is.na(df_2019$from_station_id)]
df_2019$to_station_id[is.na(df_2019$to_station_id)] <- 
  df_2019$X02...Rental.End.Station.ID[is.na(df_2019$to_station_id)]
df_2019$usertype[is.na(df_2019$usertype)] <- 
  df_2019$User.Type[is.na(df_2019$usertype)]
print("Done.")
```

#### df_2020_Q1, df_2020_Q2_to_2020_11 and df_2020_12_to_2023 are OK
```{r}
colnames(df_2017)
```

## Only select columns of interest
```{r}
df_2015 <- select(df_2015, "trip_id", "starttime", "stoptime", 
                          "bikeid", "tripduration", "from_station_id", 
                          "from_station_name", "to_station_id", 
                          "to_station_name", "usertype")

df_2016 <- select(df_2016, "trip_id", "starttime", "stoptime", 
                          "bikeid", "tripduration", "from_station_id", 
                          "from_station_name", "to_station_id", 
                          "to_station_name", "usertype")

df_2017 <- select(df_2017, "trip_id", "start_time", "end_time", 
                          "bikeid", "tripduration", "from_station_id", 
                          "from_station_name", "to_station_id", 
                          "to_station_name", "usertype")

df_2018 <- select(df_2018, "trip_id", "start_time", "end_time", 
                          "bikeid", "tripduration", "from_station_id", 
                          "from_station_name", "to_station_id", 
                          "to_station_name", "usertype")

df_2019 <- select(df_2019, "trip_id", "start_time", "end_time", 
                          "bikeid", "tripduration", "from_station_id", 
                          "from_station_name", "to_station_id", 
                          "to_station_name", "usertype")

# Attribute tripduration must be calculated
df_2020_Q1 <- select(df_2020_Q1, "ride_id", "started_at", "ended_at", 
                          "rideable_type", "start_station_id", 
                          "start_station_name", "end_station_id", 
                          "end_station_name", "member_casual")

# Attribute tripduration must be calculated
df_2020_Q2_to_2020_11 <- select(df_2020_Q2_to_2020_11, "ride_id", "started_at", "ended_at", 
                          "rideable_type", "start_station_id", 
                          "start_station_name", "end_station_id", 
                          "end_station_name", "member_casual")

# Attribute tripduration must be calculated
df_2020_12_to_2023 <- select(df_2020_12_to_2023, "ride_id", "started_at", "ended_at", 
                          "rideable_type", "start_station_id", 
                          "start_station_name", "end_station_id", 
                          "end_station_name", "member_casual")

print("Columns selection done.")
```

```{r}
colnames(df_2015)
colnames(df_2016)
colnames(df_2017)
colnames(df_2018)
colnames(df_2019)
colnames(df_2020_Q1)
colnames(df_2020_Q2_to_2020_11)
colnames(df_2020_12_to_2023)
```

```{r}
df_2015 <- rename(df_2015,
                  start_time = starttime,
                  end_time = stoptime)
df_2016 <- rename(df_2016,
                  start_time = starttime,
                  end_time = stoptime)

```


```{r}
str(df_2016)
```

### After rew renamings We create 3 main data sets as the data sets can be grouped based on the columns they contain
```{r}
df_2015_to_2019 <- bind_rows(df_2015, df_2016, df_2016, df_2017, df_2018, df_2019)

# Transform all start_station_id into characters to match data such as KA1503000044 in df_2020_12_to_2023
df_2020_Q1$start_station_id <- as.character(df_2020_Q1$start_station_id)
df_2020_Q1$end_station_id <- as.character(df_2020_Q1$end_station_id)
df_2020_Q2_to_2020_11$start_station_id <- as.character(df_2020_Q2_to_2020_11$start_station_id)
df_2020_Q2_to_2020_11$end_station_id <- as.character(df_2020_Q2_to_2020_11$end_station_id)

df_2020_to_2023 <- bind_rows(df_2020_Q1, df_2020_Q2_to_2020_11, df_2020_12_to_2023)
print("Done.")
```

Check the merging: rows should sum up
```{r}
dim(df_2015)+
  dim(df_2016)+
  dim(df_2016)+
  dim(df_2017)+
  dim(df_2018)+
  dim(df_2019)
dim(df_2015_to_2019)

dim(df_2020_Q1)+
  dim(df_2020_Q2_to_2020_11)+
  dim(df_2020_12_to_2023)
dim(df_2020_to_2023)
```
Calculate the duration of trips for df_2020_to_2023
```{r}
df_2020_to_2023$ride_length <- as.numeric(difftime(df_2020_to_2023$ended_at, 
                                        df_2020_to_2023$started_at))

# Re-organize the columns through a new selection
df_2020_to_2023 <- select(df_2020_to_2023, "ride_id", "started_at", "ended_at", 
                          "ride_length", "rideable_type", "start_station_id", 
                          "start_station_name", "end_station_id", 
                          "end_station_name", "member_casual")
head(df_2020_to_2023)
```
```{r}
str(df_2015_to_2019)
```

##### Rename columns of df_2015_to_2019
```{r}
df_2015_to_2019 <- rename(df_2015_to_2019,
                          ride_id =trip_id ,
                          started_at = start_time,
                          ended_at = end_time,
                          ride_length = tripduration,
                          rideable_type = bikeid,
                          start_station_id = from_station_id,
                          start_station_name = from_station_name,
                          end_station_id = to_station_id,
                          end_station_name = to_station_name,
                          member_casual = usertype)
print("Done.")
```

```{r}
colnames(df_2015_to_2019)
colnames(df_2020_to_2023)
```

##### Create the date, month, day, year, and week day attributes
```{r}
df_2015_to_2019$date <- as.Date(df_2015_to_2019$started_at) #The default format is yyyy-mm-dd
df_2015_to_2019$month <- format(as.Date(df_2015_to_2019$date), "%m")
df_2015_to_2019$day <- format(as.Date(df_2015_to_2019$date), "%d")
df_2015_to_2019$year <- format(as.Date(df_2015_to_2019$date), "%Y")
df_2015_to_2019$day_of_week <- format(as.Date(df_2015_to_2019$date), "%A")
print("Done with df_2015_to_2019.")

df_2020_to_2023$date <- as.Date(df_2020_to_2023$started_at) #The default format is yyyy-mm-dd
df_2020_to_2023$month <- format(as.Date(df_2020_to_2023$date), "%m")
df_2020_to_2023$day <- format(as.Date(df_2020_to_2023$date), "%d")
df_2020_to_2023$year <- format(as.Date(df_2020_to_2023$date), "%Y")
df_2020_to_2023$day_of_week <- format(as.Date(df_2020_to_2023$date), "%A")
print("Done with df_2020_to_2023")
```

```{r}
str(df_2015_to_2019)
str(df_2020_to_2023)
```
##### Summaries
```{r Summary of df_2015_to_2019}
summary(df_2015_to_2019)
```

```{r Summary of df_2020_to_2023}
summary(df_2020_to_2023)
```

##### Save the 2 formated data sets
```{r Save the 2 formated data sets}
write.csv(df_2015_to_2019, file = file.path('data', 'df_2015_to_2019.csv'), row.names = FALSE)

write.csv(df_2020_to_2023, file = file.path('data', 'df_2020_to_2023.csv'), row.names = FALSE)
```

### Next: Third step - Cleaning