---
title: "Third step: Cleaning"
author: "Amos Bationo"
date: "2023-04-18"
output: html_document
---

```{r setup, include=FALSE}
library("tidyverse")
library("lubridate")
```

# -------------- Read all extracted csv files
#______________________________________________________________________________#
### ---- Data sets
```{r Data sets, echo=FALSE}
df_2015_to_2019 <- read_csv(file.path('data', 'df_2015_to_2019.csv'))

df_2020_to_2023 <- read_csv(file.path('data', 'df_2020_to_2023.csv'))
```


## Cleaning
#### Data summaries
```{r}
dim(df_2015_to_2019)
dim(df_2020_to_2023)
```

```{r Summary of df_2015_to_2019}
summary(df_2015_to_2019)
```

```{r Summary of df_2020_to_2023}
summary(df_2020_to_2023)
```
#### Removing inconsistent data
From df_2020_to_2023, remove trips with negative or zero ride_length and where "bikes were taken out of docks and checked for quality by Divvy"
```{r}
df_2020_to_2023 <- filter(df_2020_to_2023, ride_length > 0, start_station_name != "HQ QR")
print("Done.")
```
Are there trips that have their started_at is older (greater) than their ended_at 
(even if we've already removed trips with ride_length < 0)
```{r}
nrow(filter(df_2015_to_2019, started_at > ended_at))
nrow(filter(df_2020_to_2023, started_at > ended_at))
```
Remove the 34 trips that have their started_at is older (greater) than their ended_at
```{r}
df_2015_to_2019 <- filter(df_2015_to_2019, !(started_at > ended_at))
dim(df_2015_to_2019)
```

##### The member_casual column should have only two categories: , "member" and ""casual"

```{r}
group_by(df_2015_to_2019, member_casual) %>% 
  summarise(number_of_trips = n(), avg_ride_duration = mean(ride_length))

group_by(df_2020_to_2023, member_casual) %>% 
  summarise(number_of_trips = n(), avg_ride_duration = mean(ride_length))
```
Remove the 'Dependent' category and change 'Customer' and 'Subscriber' into 'casual' and 'member' respectively
```{r}
df_2015_to_2019 <- filter(df_2015_to_2019, member_casual != 'Dependent') %>% 
  mutate(member_casual = recode(member_casual,
    'Customer' = 'casual',
    'Subscriber' = 'member'
  ))
print("Done.")
```
Check the result
```{r}
group_by(df_2015_to_2019, member_casual) %>% 
  summarise(number_of_trips = n(), avg_ride_duration = mean(ride_length))

group_by(df_2020_to_2023, member_casual) %>% 
  summarise(number_of_trips = n(), avg_ride_duration = mean(ride_length))
```
Combie=ne the two data sets
```{r}
colnames(df_2015_to_2019)
colnames(df_2020_to_2023)
```
```{r}
str(df_2015_to_2019)
str(df_2020_to_2023)
```
```{r}
# Transform ride_id into character as it is in that type in df_2015_to_2023 and contains letters there
df_2015_to_2019$ride_id <- as.character(df_2015_to_2019$ride_id)
df_2015_to_2019$rideable_type <- as.character(df_2015_to_2019$rideable_type)
df_2015_to_2019$start_station_id <- as.character(df_2015_to_2019$start_station_id)
df_2015_to_2019$end_station_id <- as.character(df_2015_to_2019$end_station_id)

df_2015_to_2023 <- bind_rows(df_2015_to_2019, df_2020_to_2023)
head(df_2015_to_2023)
```
```{r}
dim(df_2015_to_2023)
```
```{r}
str(df_2015_to_2023)
```

```{r}
dim(df_2015_to_2023)
```


```{r export cleaned df_2020_to_2023 to csv}
write.csv(df_2015_to_2023, file.path('data', 'df_2015_to_2023.csv'), row.names = FALSE)
```

### Next: Fourth step - Analysis
